---
title: "Clustering
output: html_notebook
---

```{r}
clocks <- c("McEwen", "Wu", "Hannum", "Horvath", "Levine", "Belsky", "Teschendorff", "Shireby", "Lu")
selected_clock_cpgs <- readRDS("clock_cpgs.rds")
clock_annotations <- matrix(0,
  nrow = length(selected_clock_cpgs),
  ncol = length(clocks),
  dimnames = list(selected_clock_cpgs, clocks)
)

for(j in seq_along(clocks)) {
  this_clock <- clocks[j]
  hits <- rownames(clock_annotations) %in% clock_cpgs[[ this_clock ]]
  clock_annotations[hits, this_clock] <- 1
}

correct_order <- readRDS("correct_order.rds") # is this the correct order before or after ordering??
clock_annotations <- clock_annotations[correct_order, ]

df_ann <- as.data.frame(clock_annotations, stringsAsFactors = FALSE)
df_ann[] <- lapply(df_ann, function(x) factor(x, levels = c(0,1)))

ha <- rowAnnotation(
  df = df_ann,
  col = list(
    McEwen       = c("0"="gray","1"="purple"),
    Wu           = c("0"="gray","1"="orange"),
    Hannum       = c("0"="gray","1"="red"),
    Horvath      = c("0"="gray","1"="blue"),
    Levine       = c("0"="gray","1"="green"),
    Belsky        = c("0"="gray","1"="yellow"),
    Teschendorff = c("0"="gray","1"="pink"),
    Shireby      = c("0"="gray","1"="magenta"),
    Lu           = c("0"="gray","1"="maroon")
  ),
  simple_anno_size  = unit(3, "mm"),
  show_legend       = FALSE,
  annotation_name_gp = gpar(fontsize=10, fontfamily="Helvetica")
)
```

```{r}
# params: num_clusters, SAVE, CLOCK, gender, ret, labs

# param ret: either "df" to get the matrix with clusters annotated for use in further pipeline, or "plot" to get the plot, for use in combining in Fig 2 and 5
# param labs: only relevant if ret == "plot", just specifying what labels you want

non_all <- mainPipeline(4, TRUE, FALSE, "all", "df", c("C", "D"))
clock_all <- mainPipeline(4, TRUE, TRUE, "all", "df", c("A", "B"))
male_all <- mainPipeline(4, TRUE, FALSE, "male", "df", c("A", "B"))
female_all <- mainPipeline(4, TRUE, FALSE, "female", "df", c("C", "D"))

# save for later for combination with enrichment stuff
fig2b <- non_all
fig2a <- clock_all

fig5a <- male_all
fig5b <- female_all


```

Determine enrichment of CpG sites in clusters
- enrichment of clock CpGs in clock clusters
- enrichment of aaCpGs in each window in aaCpG clusters (Fig. 3C & Fig. 3D)

*Necessary to run mainPipeline to have the matrices with cluster annotations ready, but can be done quickly by specifying save=FALSE and plot_smooth=FALSE
*Necessary to run filter code to have list1 ready

```{r}

# Run

ht <- clusterEnrichment(clock_all, 4, clock_cpgs_list)
p3 <- grid.grabExpr(draw(ht, column_title = " ", column_title_gp = gpar(fontsize=14, fontface="bold", fontfamily="Helvetica")))

ht2 <- clusterEnrichment(non_all, 4, all_list)
p4 <- grid.grabExpr(draw(ht2, column_title = " ", column_title_gp = gpar(fontsize=14, fontface="bold", fontfamily="Helvetica")))

fig2c <- plot_grid(p3, p4, labels=c("E", "F"), rel_widths=c(0.8, 1.2))
```

```{r}
# combined for fig2 -> fig2a, fig2b, p3, and p4

combined <- plot_grid(fig2a, fig2b, fig2c, nrow=3, rel_heights = c(1, 1, 0.8))
ggsave("new_figs/fig_2.pdf", plot = combined, device=cairo_pdf, 
        width = 15, height = 16, dpi=300)

# combined for fig5 <- fig5a, fig5b

combined <- plot_grid(fig5a, fig5b, nrow=2)
ggsave("new_figs/fig_5.pdf", plot = combined, device = cairo_pdf, width = 15, height = 11.42, dpi=300)


```



UNUSED BELOW

```{r}
library(circlize)
library(gridBase)

ann <- read.table("HM450.hg38.manifest.tsv", sep="\t", header=TRUE, quote="", comment.char="", fileEncoding = "UTF-8")
ann <- ann[!is.na(ann$CpG_chrm) & !is.na(ann$CpG_beg) & !is.na(ann$CpG_end), ]
rownames(ann) <- ann$probeID
#earlyagegroup <- four_cluster_select[four_cluster_select$Cluster==1 | four_cluster_select$Cluster==3, ]
#earlycpgs <- rownames(earlyagegroup)

#oldagegroup <- four_cluster_select[four_cluster_select$Cluster==2 | four_cluster_select$Cluster==4, ]
#oldcpgs <- rownames(oldagegroup)
all_aaCpGs <- readRDS("all_select_cpg.rds")
male_aaCpGs <- readRDS("male_select_cpg.rds")
female_aaCpGs <- readRDS("female_select_cpg.rds")

K450 <- ann[, 1:3]
colnames(K450) <- c("chr", "start", "end")
selected <- K450[rownames(K450) %in% all_aaCpGs, ]
maleSelected <- K450[rownames(K450) %in% male_aaCpGs, ]
femaleSelected <- K450[rownames(K450) %in% female_aaCpGs, ]
#youngSelected <- K450[rownames(K450) %in% earlycpgs, ]
#oldSelected <- K450[rownames(K450) %in% oldcpgs, ]
#all_sites <- readRDS("7_23_all_sites.rds")
#K300 <- ann[rownames(ann) %in% all_sites, 1:3]

#set.seed(123)
#M1 <- M28[sample(nrow(M28), size=1000000, replace=FALSE), ]



lgd_points = Legend(at = c("450K CpGs"), type = "points", 
    legend_gp = gpar(col = "blue"), title_position = "topleft", 
    title = " ")
lgd_points2 = Legend(at = c("Selected aaCpGs"), type = "points", 
    legend_gp = gpar(col = "red"), title_position = "topleft", 
    title = " ")
lgd_points3 = Legend(at = c("Male aaCpGs"), type = "points", 
    legend_gp = gpar(col = "green"), title_position = "topleft", 
    title = " ")
lgd_points4 = Legend(at = c("Female aaCpGs"), type = "points", 
     legend_gp = gpar(col = "purple"), title_position = "topleft", 
     title = " ")
lgd_list_vertical = packLegend(lgd_points, lgd_points2, lgd_points3, lgd_points4)

#save(M28, K300, selected, file="10_15.RData")
#load("10_15.RData")

png("new_figs/temp_circos.png", width = 8, height=6, units="in", res=300)
plot.new()
circle_size = unit(1, "snpc")
pushViewport(viewport(x = 0, y = 0.5, width = circle_size, height = circle_size,
    just = c("left", "center")))
par(omi = gridOMI(), new = TRUE)
#vps <- baseViewports()
#par(new = TRUE, fig = gridFIG())
circos.initializeWithIdeogram(chromosome.index = paste0("chr", 1:22), species="hg38")
        #circos.genomicDensity(M28, col = c("#FF000080"), track.height = 0.1)
circos.genomicDensity(K450, col=c("blue"), track.height=0.1)
circos.genomicDensity(selected, col=c("red"), track.height=0.1)
circos.genomicDensity(maleSelected, col=c("green"), track.height=0.1)
circos.genomicDensity(femaleSelected, col=c("purple"), track.height=0.1)

upViewport()
draw(lgd_list_vertical, x = circle_size, just = "left")
dev.off()
```

