---
title: "Pre-Processing"
output: html_notebook
---

```{r}
# load any extra packages needed here
# packages loaded concordantly in HelperFunctions.Rmd
```

```{r}
# Read standard CpG site list to be used (not until matrix has been generated then filtered)
ann <- read.table("HM450.hg38.manifest.tsv", sep="\t", header=TRUE, quote="", comment.char="", fileEncoding = "UTF-8")
standard_cpg_list <- ann$probeID
```


- One by one -> keep track in notebook

Start from scratch (datasets directly from OneDrive) 
- M to beta conversion ✔️
- bounding checks ✔️
- 850K to 450K platform conversion problems ✔️

   - Generate histograms of age distribution + distribution across datasets (colored + patterned)
   - generate bar plot of unique CpG distribution across datasets

- save in standardized format given by https://bio-learn.github.io/methylation-standard.html will load correctly. (slightly modified w/ row and column names)
 - format as csv
 - each row has same number of entries
 - row corresponds to measurements to a CpG site, with rownames as CpG ID
 - rownames is in standard cpg order
 - colNames has sample ID
 * no more than 1000 samples if wanted to be used with biolearn package
 - metadata must have 0 for female, 1 for male
 - column names: sample_id, age, gender, type, tissue, disease state (may need placeholder columns if datasets do not contain them)

- Made the choice to individually go through each dataset instead of writing a pipeline function to ensure that nuances do not slip through

```{r}
# 4 things to check after each dataset has been processed: metadata_tracker, running_site_list, and valid_samples, full_metadata
age_range <- 0:100
metadata_tracker <- data.frame(matrix(0, nrow=length(age_range), ncol=0))
rownames(metadata_tracker) <- paste0("age", age_range)

running_site_list <- data.frame(Count = rep(0, length(standard_cpg_list)))
rownames(running_site_list) <- standard_cpg_list

valid_samples <- list()

full_metadata <- data.frame(sample_id = NULL,
                            age = NULL, 
                            gender = NULL,
                            type = NULL,
                            tissue = NULL,
                            disease_status = NULL)

# GSE32148
# ------------------

file_path <- "original_datasets/oGSE32148_beta.xlsx"
file_path2 <- "original_datasets/oGSE32148_md.xlsx"

temp_beta <- xlsx_helper(file_path)
temp_md <- xlsx_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE32148")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

full_metadata <- rbind(full_metadata, temp_md)

# save processed data
#write.csv(new_dataset, "pp_datasets/GSE32148_pp.csv") # work-around to preserve rownames, when read using fread, rownames become the first column

# GSE36054
# ------------------

file_path <- "original_datasets/oGSE36054_beta.xlsx"
file_path2 <- "original_datasets/oGSE36054_md.xlsx"

temp_beta <- xlsx_helper(file_path)
temp_md <- xlsx_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE36054")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE36054_pp.csv")

# GSE40279
# ------------------

file_path <- "original_datasets/oGSE40279_beta.csv"
file_path2 <- "original_datasets/oGSE40279_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE40279")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE40279_pp.csv")

# GSE50660
# ------------------

file_path <- "original_datasets/oGSE50660_beta.csv"
file_path2 <- "original_datasets/oGSE50660_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE50660")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE50660_pp.csv")

# GSE50759
# ------------------

file_path <- "original_datasets/oGSE50759_beta.xlsx"
file_path2 <- "original_datasets/oGSE50759_md.xlsx"

temp_beta <- xlsx_helper(file_path)
temp_md <- xlsx_helper(file_path2)
colnames(temp_md) <- c("sample_id", "age", "gender", "type")

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

temp_beta <- m_to_b(temp_beta)
check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE50759")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE50759_pp.csv")

# GSE51057
# ------------------

file_path <- "original_datasets/oGSE51057_beta.csv"
file_path2 <- "original_datasets/oGSE51057_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE51057")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE51057_pp.csv")

# GSE53740
# ------------------

file_path <- "original_datasets/oGSE53740_beta.csv"
file_path2 <- "original_datasets/oGSE53740_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE53740")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md <- temp_md[, c(1, 2, 3, 4, 6, 5)]
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE53740_pp.csv")

# GSE61256
# ------------------

file_path <- "original_datasets/oGSE61256_beta.csv"
file_path2 <- "original_datasets/oGSE61256_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE61256")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md <- temp_md[, c(1, 2, 3, 4, 6, 5)]
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE61256_pp.csv")

# GSE67705
# ------------------

file_path <- "original_datasets/oGSE67705_beta.csv"
file_path2 <- "original_datasets/oGSE67705_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE67705")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE67705_pp.csv")

# GSE73103
# ------------------

file_path <- "original_datasets/oGSE73103_beta.csv"
file_path2 <- "original_datasets/oGSE73103_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE73103")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE73103_pp.csv")

# GSE80261
# ------------------

file_path <- "original_datasets/oGSE80261_beta.xlsx"
file_path2 <- "original_datasets/oGSE80261_md.xlsx"

temp_beta <- xlsx_helper(file_path)
temp_md <- xlsx_helper(file_path2)
colnames(temp_md)[1] <- "sample_id"

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE80261")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_subgroup <- NULL
temp_md <- temp_md[, c(1, 2, 3, 4, 6, 5)]
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE80261_pp.csv")

# GSE85568
# ------------------

file_path <- "original_datasets/oGSE85568_beta.csv"
file_path2 <- "original_datasets/oGSE85568_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)
colnames(temp_md)[1] <- "sample_id"

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

temp_beta <- limit_bounds(temp_beta)
check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE85568")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md <- temp_md[, c(1, 2, 3, 4, 6, 5)]
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE85568_pp.csv")

# GSE89253
# ------------------

file_path <- "original_datasets/oGSE89253_beta.csv"
file_path2 <- "original_datasets/oGSE89253_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)
colnames(temp_md)[1] <- "sample_id"

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

temp_beta <- m_to_b(temp_beta)
check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE89253")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE89253_pp.csv")

# GSE90124
# ------------------

file_path <- "original_datasets/oGSE90124_beta.csv"
file_path2 <- "original_datasets/oGSE90124_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE90124")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE90124_pp.csv")

# GSE94734
# ------------------

file_path <- "original_datasets/oGSE94734_beta.xlsx"
file_path2 <- "original_datasets/oGSE94734_md.xlsx"

temp_beta <- xlsx_helper(file_path)
temp_md <- xlsx_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE94734")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE94734_pp.csv")

# GSE106648
# ------------------

file_path <- "original_datasets/oGSE106648_beta.csv"
file_path2 <- "original_datasets/oGSE106648_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE106648")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md <- temp_md[, c(1, 2, 3, 4, 6, 5)]
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE106648_pp.csv")

# GSE114134
# ------------------

file_path <- "original_datasets/oGSE114134_beta.csv"
file_path2 <- "original_datasets/oGSE114134_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

temp_beta <- as.matrix(temp_beta)
temp_beta <- liftover(temp_beta)

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE114134")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md <- temp_md[, c(1, 2, 3, 4, 6, 5)]
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE114134_pp.csv")

# GSE124076
# ------------------
# CUT FOR NOW

# GSE124366
# ------------------

file_path <- "original_datasets/oGSE124366_beta.csv"
file_path2 <- "original_datasets/oGSE124366_md.csv"

temp_beta <- csv_helper(file_path)
temp_md <- csv_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE124366")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$tissue <- "none"
temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE124366_pp.csv")

# GSE137495
# ------------------

file_path <- "original_datasets/oGSE137495_beta.xlsx"
file_path2 <- "original_datasets/oGSE137495_md.xlsx"

temp_beta <- xlsx_helper(file_path)
temp_md <- xlsx_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE137495")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE137495_pp.csv")

# GSE138279
# ------------------

file_path <- "original_datasets/oGSE138279_beta.xlsx"
file_path2 <- "original_datasets/oGSE138279_md.xlsx"

temp_beta <- xlsx_helper(file_path)
temp_md <- xlsx_helper(file_path2)

print(paste0("Total ages NA: ", sum(is.na(as.numeric(temp_md$age)))))
print(paste0("Ages above 80: ", sum(as.numeric(temp_md$age)>80, na.rm=TRUE)))

temp_beta <- as.data.frame(temp_beta)
rownames(temp_beta) <- temp_beta[, 1]
temp_beta <- temp_beta[, -1]

check_bounds(temp_beta)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(temp_beta)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(temp_beta)
new_dataset[rownames(temp_beta), ] <- temp_beta

metadata_tracker <- updateMetadata(metadata_tracker, temp_md, "GSE138279")
valid_samples <- updateValidSamples(valid_samples, temp_md, 80)
running_site_list[rownames(temp_beta), ] <- running_site_list[rownames(temp_beta), ] + 1

temp_md$disease_status <- "none"
colnames(temp_md) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")
full_metadata <- rbind(full_metadata, temp_md)

#write.csv(new_dataset, "pp_datasets/GSE138279_pp.csv")

# Working with new datasets (not on GEO, not from Hari or Wenshu)

library(GEOquery)

#GSE62924
# ------------------

gse <- getGEO("GSE62924", GSEMatrix = TRUE)
gse_info <- gse[[1]]
betas <- exprs(gse_info)
metadata <- pData(gse_info)

metadata <- metadata[, c(2, 8)]
metadata$gender <- "none"
metadata$type <- "genomic"
metadata$disease_state <- "none"
metadata$age <- 0
metadata <- metadata[, c(1, 6, 3, 4, 2, 5)]
colnames(metadata) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")

check_bounds(betas)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(betas)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(betas)
new_dataset[rownames(betas), ] <- betas

metadata_tracker <- updateMetadata(metadata_tracker, metadata, "GSE62924")
valid_samples <- updateValidSamples(valid_samples, metadata, 80)
running_site_list[rownames(betas), ] <- running_site_list[rownames(betas), ] + 1

full_metadata <- rbind(full_metadata, metadata)

#write.csv(new_dataset, "pp_datasets/GSE62924_pp.csv")

#GSE51180
# ------------------

gse <- getGEO("GSE51180", GSEMatrix = TRUE)
gse_info <- gse[[1]]
betas <- exprs(gse_info)
metadata <- pData(gse_info)

metadata <- metadata[, c(1, 2, 6, 37)]
metadata$tissue <- "blood spot"
metadata$disease_status <- "none"
metadata$age <- ifelse(grepl("at 18 years age", metadata[, 1]), 18, 0)
metadata <- metadata[, -1]
metadata <- metadata[, c(1, 6, 3, 2, 4, 5)]
colnames(metadata) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")

check_bounds(betas)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(betas)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(betas)
new_dataset[rownames(betas), ] <- betas

metadata_tracker <- updateMetadata(metadata_tracker, metadata, "GSE51180")
valid_samples <- updateValidSamples(valid_samples, metadata, 80)
running_site_list[rownames(betas), ] <- running_site_list[rownames(betas), ] + 1

full_metadata <- rbind(full_metadata, metadata)

#write.csv(new_dataset, "pp_datasets/GSE51180_pp.csv")

#GSE30870
# ------------------

gse <- getGEO("GSE30870", GSEMatrix = TRUE)
gse_info <- gse[[1]]
betas <- exprs(gse_info)
metadata <- pData(gse_info)

metadata <- metadata[, c(2, 6, 8, 12)]
metadata$age <- ifelse(metadata[, 3] == "Newborns", 0, 100) # not all nonagenarians are 100, just setting it do that because it will get cancelled out anyway
metadata$disease_status <- "none"
metadata <- metadata[, -3]
metadata$gender <- "none"
metadata <- metadata[, c(1, 4, 6, 2, 3, 5)]
colnames(metadata) <- c("sample_id", "age", "gender", "type", "tissue", "disease_status")

print(paste0("Ages above 80: ", sum(as.numeric(metadata$age)>80, na.rm=TRUE)))

check_bounds(betas)

new_dataset <- data.frame(matrix(nrow=length(standard_cpg_list), ncol=ncol(betas)), row.names = standard_cpg_list)
colnames(new_dataset) <- colnames(betas)
new_dataset[rownames(betas), ] <- betas

metadata_tracker <- updateMetadata(metadata_tracker, metadata, "GSE30870")
valid_samples <- updateValidSamples(valid_samples, metadata, 80)
running_site_list[rownames(betas), ] <- running_site_list[rownames(betas), ] + 1

full_metadata <- rbind(full_metadata, metadata)

#write.csv(new_dataset, "pp_datasets/GSE30870_pp.csv")

full_metadata$age <- round(as.numeric(full_metadata$age))

#temp <- unlist(valid_samples)
#saveRDS(temp, "valid_samples.rds")
#saveRDS(full_metadata, "full_metadata.rds")
```

Generate age distribution figures

```{r}
# from last chunk: metadata_tracker & running_site_list

library(RColorBrewer)
library(ggplot2)
library(ggpubr)

coul <- colorRampPalette(brewer.pal(8, "Set3"))(20)

data_percentage <- apply(metadata_tracker, 1, function(x) {x*100/sum(x, na.rm=T)})

barplot(data_percentage[, 1:81], col=coul, border="white", xlab=NULL, axisnames=FALSE)

temp <- metadata_tracker
plot_data <- data.frame(ages = 0:80, count = rowSums(temp[1:81, ]))

ggbarplot(plot_data, x = "ages", y = "count")
```

Get list of valid sites for further analysis

```{r}
temp <- running_site_list[na.omit(running_site_list)$Count>=18, , drop=FALSE]
cpg_list <- rownames(temp)

# Filter again for some reason.. some issues with prev subsetting
cpg_list <- intersect(cpg_list, standard_cpg_list)

#saveRDS(cpg_list, "cpg_list.rds")
```

Extract gendered samples

```{r}
full_metadata <- readRDS("full_metadata.rds")

saveRDS(male_samples, "male_samples.rds")
saveRDS(female_samples, "female_samples.rds")
```


Get clock sites
```{r}
library(readxl)
library(DunedinPACE)

Hannum_CpGs <- read_excel("CpGsToInvestigate/hannum_cpgs.xlsx")
Hannum_CpGs <- Hannum_CpGs$Marker #71
Levine_CpGs <- read.csv("CpGsToInvestigate/levine_cpgs.csv", stringsAsFactors=FALSE)
Levine_CpGs <- Levine_CpGs[-1, ]
Levine_CpGs <- Levine_CpGs$CpG #513
Horvath_CpGs <- read.csv("CpGsToInvestigate/horvath_cpgs.csv", stringsAsFactors=FALSE)
Horvath_CpGs <- Horvath_CpGs[-(1:3), 1, drop=FALSE]
Horvath_CpGs <- Horvath_CpGs[, 1] #353
McEwen_CpGs <- read.csv("CpGsToInvestigate/mcewen_cpgs.csv")
McEwen_CpGs <- McEwen_CpGs$CPG #94
Wu_CpGs <- read_excel("CpGsToInvestigate/aging-11-102399-s003..xlsx")
Wu_CpGs <- Wu_CpGs[-1, ]
Wu_CpGs <- Wu_CpGs$CpGs #111
Belsky_CpGs <- getRequiredProbes(backgroundList=FALSE)
Belsky_CpGs <- unlist(Belsky_CpGs) #173
load("CpGsToInvestigate/epitoc.Rd")
Teschendorff_CpGs <- as.data.frame(dataETOC2.l[1])
Teschendorff_CpGs <- rownames(Teschendorff_CpGs) #163
Shireby_CpGs <- readLines("CpGsToInvestigate/CorticalClockCoefs.txt")[-1]
Shireby_CpGs <- sapply(strsplit(Shireby_CpGs, " "), `[`, 1)
Shireby_CpGs <- unlist(Shireby_CpGs) #347
Lu2_CpGs <- read_excel("CpGsToInvestigate/lu2.xlsx")
Lu2_CpGs <- rbind(colnames(Lu2_CpGs), Lu2_CpGs)
Lu2_CpGs <- Lu2_CpGs[, 1]
Lu2_CpGs <- as.character(Lu2_CpGs[[1]]) #140
clock_cpgs <- list(
  McEwen = McEwen_CpGs,
  Wu = Wu_CpGs,
  Hannum = Hannum_CpGs,
  Horvath = Horvath_CpGs,
  Levine = Levine_CpGs,
  Belsky = Belsky_CpGs,
  Teschendorff = Teschendorff_CpGs,
  Shireby = Shireby_CpGs,
  Lu = Lu2_CpGs
)

cpgs <- unlist(clock_cpgs)
cpgs <- unique(cpgs) #1868 total sites

print(paste0("Total Unique CpG sites from 9 clocks explored: ", length(cpgs)))

filter_cpgs <- intersect(cpgs, cpg_list)
#saveRDS(filter_cpgs, "clock_cpgs.rds")

# UpSet plot for clock site overlap 
# Will modify aesthetics to match when aesthetics for previous plots are finalized.
#m1 <- make_comb_mat(clock_cpgs, mode="distinct")
#UpSet(m1, set_order = rownames(m1))
#dev.copy(png, "figs/clockcpgupset.png", width=1800, height=900, res=300)
#dev.off()
```

Make Splits

```{r}
cpg_list <- readRDS("cpg_list.rds")
full_metadata <- readRDS("full_metadata.rds")
valid_samples <- readRDS("valid_samples.rds")
output_folder <- "splits/"

# makeSplits(cpg_list, full_metadata, valid_samples, 50000, output_folder)
```

Make matrix from splits

```{r}
cpg_list <- readRDS("cpg_list.rds")
full_metadata <- readRDS("full_metadata.rds")
valid_samples <- readRDS("valid_samples.rds")
output_folder <- "splits/"

male_samples <- full_metadata[full_metadata$gender=="male" | full_metadata$gender=="Male", ]$sample_id
male_samples <- intersect(male_samples, valid_samples)
female_samples <- full_metadata[full_metadata$gender=="female" | full_metadata$gender=="Female", ]$sample_id
female_samples <- intersect(female_samples, valid_samples)

# generateMatrix(input_folder, cpg_list, full_metadata, gender, average, output_folder)

#generateMatrix("splits/", cpg_list, full_metadata, "all", TRUE, "matrix/")
#generateMatrix("splits/", cpg_list, full_metadata, "all", FALSE, "matrix/")
generateMatrix("splits/", cpg_list, full_metadata, "male", TRUE, "matrix/")
generateMatrix("splits/", cpg_list, full_metadata, "male", FALSE, "matrix/")
generateMatrix("splits/", cpg_list, full_metadata, "female", TRUE, "matrix/")
generateMatrix("splits/", cpg_list, full_metadata, "female", FALSE, "matrix/")

```
